Starcode: an exact algorithm for sequence clustering

### 2014-07-05 doc/document.tex
Manuscript ready for submission.

\begin{abstract}
\section{Motivation:}
The increasing throughput of sequencing technologies offers new applications
and challenges for computational biology. One such application is the use of
random barcodes to trace and quantify transcripts or lineages in experimentals
setups. The high error rate of modern sequencers calls for additional
post-processing techniques capable of detecting and reverting the misreads.
However, in the absence of a reference population, the problem amounts to
performing a pairwise comparison of all the barcodes, which is unfeasible for
excessive computational complexity.

\section{Results:}
Here we address this problem and describe an exact algorithm to determine which
pairs of sequences lie within a given Levenshtein distance. The matched pairs
are merged into clusters represented by a canonical sequence. The effiency of
starcode is attributable to the poucet search, a novel  implementation of the
Needleman-Wunsch algorithm performed on the nodes of a trie. Parallelization
achieves linear scaling of performance on multi-core machines. On the task of
clustering random barcodes, starcode outperforms short read mappers and sequence
clustering algorithms in both speed and precision.

@@

Sequencing errors can be discovered by mapping the reads onto a reference, if it
is available. When the sequences are random or drawn from an unknown
reference, clustering is the best option to tell real from spurious reads. One
such case is the use of random barcodes to track cells or transcripts
\citep{pmid18809713, pmid23953119}. Sequencing errors will create erroneous
barcodes that have to be reverted to the original sequence. Hands on experience
with real datasets show that this step becomes limiting when
the number of unique sequences is high. In search for a solution to this problem, we
realized that heuristic approaches rely on assumptions that may not hold
as technologies evolve. We therefore set out to find an exact algorithm.

@@

Here we describe the starcode algorithm and we benchmark it against existing
software that perform related tasks. We show that starcode is both faster and
more precise than the alternatives, achieving perfect clustering on ideal datasets.
We further show that starcode is a good candidate to cluster k-mers from large
datasets from bacterial or even metazoan genomes.

@@

\begin{figure}[!tpb]
\centerline{\includegraphics{NW.pdf}}
\caption{Needleman-Wunsch (NW) alignment. \textbf{a}
  Alignment of GTTGCA and GATCCA using
the NW alignment matrix. The margins (purple) are initialized and the cells of
the matrix (yellow) are computed from left to right and from top to bottom by
the following
dynamic programming algorithm. The value in each cell is computed from the values
of the neighboring top, left and top-left cells. Denoting $a$, $b$ and $c$ these
respective terms, the value in the cell of coordinates $(i,j)$ is computed as
$\min(a+1, b+1, c)$ if the $i$-th letter from the first sequence is identical to
the $j$-th letter from the second, or $\min(a+1, b+1, c+1)$ if the letters are
different. The Levenshtein distance
between the two sequences is found in the bottom right cell. \textbf{b}
Lower complexity algorithm to
determine whether GTTGCA and GATCCA are 2-neighbors. The values in purple cells
are set during initialization. The dynamic programming algorithm proceeds as above,
with the difference that it is interrupted if the value of a diagonal cell (bold
borders) is larger than 2. The values in purple cells are not always the same as
their equivalent in a (purple arrow), but the values in the yellow cells are
nevertheless identical. The values of the white cells are never computed, which
contributes to making the algorithm less complex.}\label{fig:NW}
\end{figure}

@@

This method can be used for inexact matching of sequences against a reference
set indexed as prefix tree, also known as a trie \citep{ukkonen}. The terms of the
matrix are updated row-wise as a depth-first search traverses the trie from the
root, as illustrated in Figure~\ref{fig:trie}. Every time a node is visited, the row
corresponding to its depth is recomputed. If the threshold value $d$ is exceeded
for a diagonal term, the Levenshtein distance for all the downstream sequences is
also necessarily greater than $d$. Therefore, no more hits are to be discovered
in this path and the depth-first search backtracks to the parent node. When the
process halts, every tail node
(corresponding to a sequence of the database) on the path of this search is a
$d$-neighbor of the query. This method is efficient because it eliminates large
areas of the search space, and because the NW alignment of the query with each
prefix of the database is computed only once.

\begin{figure}[!tpb]
\centerline{\includegraphics{trie.pdf}}
\caption{Trie query with NW alignment. Each sequence of the index is a path in
the trie. The query GTTGCA is written at the top of a NW matrix, which is
initialized as shown on Figure~\ref{fig:NW}b. The trie is traversed by a
depth-first search (red path) from the root. At each depth, the node added to
the path is written on the left of the NW matrix and the row is computed.
Checkpoints from 1 to 4 (circled red numbers) show the state of the NW matrix
as the search proceeds. The node labelled 3 is a leaf and thus corresponds to a
2-neighbor of the query. The search path then backtracks to the node labelled 2
and the last rows of the NW are erased. The search path then goes to node labelled
4, in which case the newly computed diagonal cell exceeds the threshold (circled
in red). Even if this node has children, they are not visited (red crosses).}
\label{fig:trie}
\end{figure}

@@

Since two exact words at different positions in the sequence shall not
represent a lookup match, we store and query the words against $k+1$
distinct lookup tables. In order to allow insertions and deletions, 
the lookup search must also query shifted versions of each word. In
the worst matching case, only one word is found and the mismatches are
evenly distributed in the query sequence. In such case, the maximum 
number of cumulative insertions/deletions at the $i$-th word is
$i-1$. Hence, the $i$-th word has to be shifted and queried against
the lookup table $2(i-1)+1$ times to cover all the possible
cases. Hence, the last word will be queried $2\tau+1$ times, including
$\tau$ shifts on the right and $\tau$ shifts on the left. For the
latter shift we make use of the $\tau$ nucleotides stored in the
tail. The total number of lookups per sequence for a given $\tau$ is
$(\tau+1)^2$.

@@

Figure~\ref{fig:perf}a shows the running time of starcode as a
function of the number of input sequences $n$. In double logarithmic scale
the trend is a straight line with slope 1.7, which suggests that the
running time complexity of starcode is approximately $O(n^{1.7})$.
Figure~\ref{fig:perf}b shows that the running time grows exponentially
as a function of the maximum Levenshtein distance used for clustering.
This is a common feature of exact algorithms based on trie search. As
this parameter increases, the poucet search bails out at a deeper
average depth in a trie that fans out exponentially.
The running time first increases as a function of the sequence length,
but then plummets and stays low (Figure~\ref{fig:perf}c).
Beyond a certain length, the lookup search algorithm starts to be
efficient, and most of the queries are resolved without the need for
trie-search.
Finally, we show the scalability of starcode with increasing
number of treahds in Figure~\ref{fig:perf}d. The search algorithm is fully
parallel and the relative performance increases linearly up to 8 threads.
The bending observed thereafter has two sources. The first is 
that the input reading and clustering steps are brief but not parallel,
the second is due to hardware limitations, i.e. there is insufficient
memory bandwidth to satisfy the increased demand of memory accesses.

@@

We generated 1 million random 40-mers duplicated 47 times. For each
40-mer, we also inserted 3 addional mutant sequences to the pool
by sampling again 3 nucleotides at random positions. The probability
that such a mutant sequence is within a distance 3 of any other
40-mer is of the order of $10^{-12}$ and can be neglected. Each
cluster thus consists of 50 sequences, of which 3 have a Levenshtein
distance 3 from the centroid of the cluster (the canonical
representant).

All software were set to run on a 6-core single-processor Intel Xeon
E5-2620 system with 50~GB of DDR3-RAM at 1333~Mhz. The execution
options for each software are summarized in
Table~\ref{tab:soft-options}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%             WHERE TO PUT THIS NOW??                               %
%Denoted as \emph{sequential R} is the
%sequential algorithm of \cite{pmid23953119} which is implemented in R
%and relies on the \texttt{stringdist} function of the
%\texttt{stringdist} package.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{table}[h]
\centering
\caption{Software execution parameters}
\resizebox{\columnwidth}{!}{
\begin{tabular}{|l|l|}
    \hline
    starcode & \texttt{-t 20}\\ \hline
    CD-HIT 4.6.1 & \texttt{-c 0.925 -T 20 -M 0}\\ \hline
    USEARCH 7.0.1090  & \texttt{-threads 20 -cluster\_fast -id 0.925}\\ \hline
        Bowtie2 indexer & \texttt{bowtie2-build --quiet}\\ \hline
    Bowtie2 2.1.0 & \texttt{-f -a -p 20 --no-hd --very-sensitive}\\ \hline
        BWA indexer & \texttt{bwa index -a is}\\ \hline
    BWA 0.7.9a & \texttt{mem -t 20 -a -k1 -B0} \\ \hline
    GEM indexer & \texttt{gem-indexer -T 12} \\ \hline
    GEM 1.423 & \texttt{-e3 -s4 -T12 --granularity 100000}\\ \hline
    \emph{sequential R} & \texttt{method='lv', maxDist=2}\\ \hline
\end{tabular}
}
\label{tab:soft-options}
\end{table}

Using 12 cores, starcode clustered the shuffled 50 million sequences
in 5 minutes and 3 seconds without a single error (the output
consisted of 1 million clusters of size 50). There is no obvious
way to compare exact algorithms with heuristic algorithms such as
CD-HIT and USEARCH. We decided to allocate the heuristics 6 minutes
on the same dataset ($\approx$ 20\% additional time) and use the
parameters giving the best accurary under this constraint.
However, we could not find a combination of
parameters such that either of them could terminate in less than
6 minutes. To simplify the task, we extracted the unique sequences
from the input file, bringing it down to 4,000,000 and followed
the same rationale. We could not find any set of parameters such
that USEARCH could process this input in less than 6 minutes,
but CD-HIT could perform the task when run as shown in
Table~\ref{tab:soft-options}.

we decided to 
allocating them 6 minutes with the same computer resources 
USEARCH could not be
run on the same input because
The running times
of CD-HIT and USEARCH were [59] and [64] minutes respectively, and they
identified 2,362,352 and 2,739,954 clusters respectively. The results
are summarized in Figure~\ref{fig:benchmark}. Both algorithms rely on
heuristics, explaining why the number of clusters is off target. These
results clearly demonstrate that starcode outperforms other clustering
algorithms both in speed and accuracy with random input sequences.

We also compared starcode on the biological benchmark dataset proposed
by \cite{pmid23060610}.

\begin{figure}[!tpb]
\centerline{\includegraphics{benchmark.pdf}}
\caption{Benchmark of starcode against sequence clustering algorithms
and short read mappers. \textbf{a} Running time compared to CD-HIT and
USEARCH on the dataset described in the text. \textbf{b} Error rate
compared to CD-HIT and USEARCH, expressed as the percentage of clusters
failing to be merged. \textbf{c} Running time compared to Bowtie2, BWA
and GEM on this dataset. The running time is decomposed in extraction
of unique sequences (purple), Burrows-Wheeler indexing (yellow), and
query proper (white). The extraction was performed with the Linux sort
command, so the time is the same in every case. \textbf{d} Error rate
compared to Bowtie2, BWA and GEM, expressed as the proportion of matches
with a single hit. This is an underestimate of the true error rate.}
\label{fig:benchmark}
\end{figure}

\subsection{Benchmark against short read mappers}
A potential strategy to cluster short sequences is to use last generation short
read mappers to match a set of sequences against itself. Once matches are
available, several community detection algorithms can be used to identify the
clusters. We benchmarked starcode against the short read mappers Bowtie2
\citep{pmid22388286}, BWA \citep{pmid19451168} and GEM \citep{pmid23103880}.
Because short read
mappers do not perform clustering, we only evaluated their performance on the
matching problem.

For this test, we used the same dataset described above, in which every
sequence has at least 2 distinct matches at a distance 3 (mutant sequences
from the same cluster may be at a distance up to 6 from each other). This
means that we
can use matches with a single hit (themselves) as a lower bound on the inaccuracy.
We used unique sequences to build an index and queried the index with the same
sequences. Figure~\ref{fig:benchmark}c shows the running time of starcode along with the
decomposition of the running time for the three mappers. Note that the time to
extract unique sequences is always the same because we used the Linux
\texttt{sort -u} command in every case. The running time of
starcode is slightly lower than short read mappers when all steps ares
taken into account.  Figure~\ref{fig:benchmark}d shows the lower bound on
the error rate, as estimated by the proportion of single hits. BWA was
found to be the most accurate aligner with a lower bound of 31\% sequences.
with a single match. This measure is an underestimate of the true error
rate, but given that starcode is an exact algorithm, a more accurate
measure is not necessary. In conclusion, starcode can compete with the
short read aligners on the matching problem, and it provides a precision
that none of the other tools can offer.

@@

We tested the efficiency of starcode on the TRIP dataset from
\cite{pmid23953119}. In the experiment labelled mPGKA, we identified
24.1 million (91\%) barcode-containing reads out of 26.6 million, consisting
of approximately 223,000 and 220,000 unique barcode sequences for PCR replicates
1 and 2, respectively. Following the authors, we kept only the barcodes with
at least 5 counts and performed clustering as described: ``First we sorted
barcodes according to their counts. Then, for each barcode (starting from the
most frequent one), we identified and removed all its mutant versions, defined
as barcodes within a Hamming distance of 2.'' This method is here on referred
to as the ``sequential algorithm''. When replacing the Hamming distance by the
Levenshtein distance, starcode produced exactly the same output as the sequential
algorithm. The running time of starcode averaged over the replicates was
2.90 seconds, \textit{versus} more than 3 minutes 40 seconds, which represents a
75-fold speedup.

@@

We set up a test based on the meningitis-causing agent
\textit{Neisseria meningitidis}. The genome of this bacterium is
interspersed with a frequent 12 bp sequence known as DNA uptake
sequence \citep{pmid10673000}. We extracted the 12-mers from both
orientations of  the 2.19 Mb genome, yielding 4.39 million 12-mers,
consisting of 2.77 million unique sequences. Clustering the 12-mers
with starcode within a Levenshtein distance of 2 took less than 45
seconds with 12 cores. We identified the known DNA uptake sequence
of \textit{Neisseria meningitidis} (ATGCCGTCTGAA) as the most abundant
12-mer, with 1466 exact and 2096 inexact hits. This result testifies
to the fact that starcode can be used to identify biologically
relevant motifs in bacterial genomes.

To test starcode on another application, we used the RNA-protein
interaction data produced by RNAcompete \citep{pmid19561594}. The mammalian
splicing factor SRSF1 is known to bind RNA GA-rich motifs, but there is
some disagreement about the motif that it recognizes \citep{pmid23562324}.
For each SRSF1 replicate of the RNAcompete dataset,
we replaced the RNAcompete signals by their rank and
extracted the 10-mers from the microarray probes. The 10-mers were given
a score equal to the rank of the probe they belong, and enriched motifs
were found using the sphere clustering of starcode with maximum
Levenshtein distance 2. The score of the most enriched 10-mer is thus the
sum of the ranks of all 10-mers within this distance. Clustering the 6.3
million extracted 10-mers with 12 cores took about 20 seconds for each
replicate.  The most enriched 10-mers were AGGACACGGA, AGGACACGGA,
AGGACGGAGG, AGGACGGAGG, AGGACACGGA and AGGATACAGG. Except for the last
replicate, the motifs consist of AGGAC and GGA, with a spacer of variable
length. This suggests that the binding of SRSF1 to RNA may be flexible,
which would explain the disagreement between the motifs derived from 6-mers
or 7-mers.

\section{Discussion and conclusion}
Through the parallel poucet search algorithm, starcode implements an exact
sequence clustering algorithm that is even faster than widely used heuristics.
By design, starcode is tailored to process high throughput sequencing data on
multi-core platforms. Our benchmark shows that starcode is superior in speed
and accuracy compared to other methods on the problem of clustering short random
sequences. Part of the reason is that such task is stretching the tested
software far from their initial design. Sequence clustering algorithms are
meant to cluster long biological sequences, while short read mappers are
meant to map reads on potentially large genomes. In this respect, starcode
fills a need arising from the development of barcoding technologies.

The speed of starcode also makes it proficient at more general clustering
problems, such as identifying enriched k-mers in genomes and in experimental
data. Here we have given two examples of such applications. In the first, we
recover a known enriched 12-mer in the genome of \textit{Neisseria meningitidis}.

The current version of starcode has been mildly optimized for memory
consumption. The memory footprint depends on the number of sequences to
cluster (because the sequences of the input set are loaded in memory as a
group of tries) and on the mean number of matches per sequence. Every match
has to be stored until the clustering phase, which may represent a large
memory overhead. As counterintuitive as it may seem, long queries will usually
impose a lower memory footprint because the matches between sequences are less
frequent.

The speed of starcode could also benefit other applications, such as searching
shared subsequences during genome assembly. However, starcode supports only global
and not local alignment. A workaround is to compare k-mers as we have done
in the examples above, but this approach is expected to be memory demanding.
Even though extending the starcode algorithm to support clustering of protein
sequences is straightforward, it is not expected to meet the performance level
of the presented algorithm. On the one hand, there are more letters in the protein
alphabet, so the shared prefix between consecutive queries is inherently
shorter on average, which defeats the poucet search. On the other hand, if
a threshold distance of the order of 1 per 10 nucleotides is sufficient to
identify similar DNA sequences, a distance of 1 per 10 residues is
unrealistically low for protein sequences, and this parameter critically
determines the speed of starcode. More generally, storing computational
intermediates for shared prefixes could find some applications in other
algorithms such as short read mappers. The idea of the poucet search seems
simple in retrospect, but it is a powerful way to tap into the data
structuration provided by string sorting.

@@

\section*{Acknowledgement}
We would like to thank...

>>>>

\begin{abstract}
\section{Motivation:}
The increasing throughput of sequencing technologies offers new
applications and challenges for computational biology. One such
application is the use of random barcodes to trace and quantify
transcripts or lineages in experimental setups. The high error
rate of modern sequencers calls for additional post-processing
techniques capable of detecting and reverting the misreads.
However, in the absence of a reference population, the problem
amounts to performing a pairwise comparison of all the barcodes,
which is unfeasible for excessive computationally complexity.

\section{Results:}
Here we address this problem and describe an exact algorithm to
determine which pairs of sequences lie within a given Levenshtein
distance. The matched pairs are merged into clusters represented
by a canonical sequence. The effiency of starcode is attributable
to the poucet search, a novel  implementation of the Needleman-Wunsch
algorithm performed on the nodes of a trie. On the task of clustering
random barcodes, starcode outperforms sequence clustering algorithms
in both speed and precision. We further show that starcode can also
be used to identify enriched motifs in DNA and RNA sequences.

@@

Sequencing errors can be discovered by mapping the reads onto a reference, if it
is available. When the sequences are random or drawn from an unknown
reference, clustering is the best option to tell real from spurious reads. One
such case is the use of random barcodes to track cells or transcripts
\citep{pmid18809713, pmid23953119}. Sequencing errors will create erroneous
barcodes that have to be reverted to the original sequence. Hands on
experience with real datasets shows that this step becomes limiting when
the number of unique sequences is high. In search for a solution to this problem, we
realized that heuristic approaches rely on assumptions that may not hold
as technologies evolve. We therefore set out to find an exact algorithm.

@@

Here we describe the starcode algorithm and we benchmark it against
existing software that perform related tasks. We show that starcode is
both faster and more precise than the alternatives, achieving perfect
clustering on ideal datasets. We further show that starcode can be used
to identify enriched motifs in large datasets such as bacterial genomes
or protein-DNA/protein-RNA interaction experiments.

@@

\begin{figure}[!tpb]
\centerline{\includegraphics{NW.pdf}}
\caption{Needleman-Wunsch (NW) alignment. \textbf{a}
Alignment of GTTGCA and GATCCA using the NW alignment matrix. The
margins (purple) are initialized and the cells of the matrix (yellow)
are computed from left to right and from top to bottom by the following
dynamic programming algorithm. The value in each cell is computed from
the values of the neighboring top, left and top-left cells. Denoting
$a$, $b$ and $c$ these respective terms, the value in the cell of
coordinates $(i,j)$ is computed as $\min(a+1, b+1, c)$ if the $i$-th
letter from the first sequence is identical to the $j$-th letter from
the second, or $\min(a+1, b+1, c+1)$ if the letters are different. The
Levenshtein distance between the two sequences is found in the bottom
right cell. \textbf{b} Lower complexity algorithm to determine whether
GTTGCA and GATCCA are 2-neighbors. The values in purple cells are set
during initialization. The dynamic programming algorithm proceeds as
above, with the difference that it is interrupted if the value of a
diagonal cell (bold borders) is larger than 2. The values in purple
cells are not always the same as their equivalent in \textbf{a} (purple
arrow), but the values in the yellow cells are nevertheless identical.
The values of the white cells are never computed, which contributes to
making the algorithm less complex.}\label{fig:NW}
\end{figure}

@@

This method can be used for inexact matching of sequences against a
reference set indexed as a prefix tree, also known as a trie
\citep{ukkonen}. The terms of the matrix are updated row-wise as a
depth-first search traverses the trie from the root, as illustrated
in Figure~\ref{fig:trie}. Every time a node is visited, the row
corresponding to its depth is recomputed. If the threshold value $d$
is exceeded for a diagonal term, the Levenshtein distance for all the
downstream sequences is also necessarily greater than $d$. Therefore,
no more hits are to be discovered in this path and the depth-first
search backtracks to the parent node. When the process halts, every
tail node (corresponding to a sequence of the database) on the path
of this search is a $d$-neighbor of the query. This method is
efficient because it eliminates large areas of the search space, and
because the NW alignment of the query with each prefix of the database
is computed only once.

\begin{figure}[!tpb]
\centerline{\includegraphics{trie.pdf}}
\caption{Trie query with NW alignment. Each sequence of the index is a
path in the trie. The query GTTGCA is written at the top of a NW matrix,
which is initialized as shown on Figure~\ref{fig:NW}b. The trie is
traversed by a depth-first search (red path) from the root. At each
depth, the node added to the path is written on the left of the NW
matrix and the row is computed. Checkpoints from 1 to 4 (circled red
numbers) show the state of the NW matrix as the search proceeds. The
node labeled 3 is a leaf and thus corresponds to a 2-neighbor of the
query. The search path then backtracks to the node labeled 2
and the last rows of the NW are erased. The search path then goes to
the node labeled 4, in which case the newly computed diagonal cell
exceeds the threshold (circled in red). Even if this node has children,
they are not visited (red crosses).}
\label{fig:trie}
\end{figure}

@@

Since two exact words at different positions in the sequence shall not
represent a lookup match, we store and query the words against $k+1$
distinct lookup tables. In order to allow insertions and deletions, 
the lookup search must also query shifted versions of each word. In
the worst matching case, only one word is found and the mismatches are
evenly distributed in the query sequence. In such case, the maximum 
number of cumulative insertions/deletions at the $i$-th word is
$i-1$. Hence, the $i$-th word has to be shifted and queried against
the lookup table $2(i-1)+1$ times to cover all the possible
cases. The last word will be queried $2\tau+1$ times, including
$\tau$ shifts on the right and $\tau$ shifts on the left. For the
latter shift we make use of the $\tau$ nucleotides stored in the
tail. The total number of lookups per sequence is $(\tau+1)^2$.

@@

Figure~\ref{fig:perf}a shows the running time of starcode as a
function of the number of input sequences $n$. In double logarithmic scale
the trend is a straight line with slope 1.7, which suggests that the
running time complexity of starcode is approximately $O(n^{1.7})$.
Figure~\ref{fig:perf}b shows that the running time grows exponentially
as a function of the maximum Levenshtein distance used for clustering.
This is a common feature of exact algorithms based on trie search. As
this parameter increases, the poucet search bails out at a deeper
average depth in a trie that fans out exponentially.
As a function of the sequence length, the running time first increases 
but then plummets and stays low (Figure~\ref{fig:perf}c).
Beyond a certain length, the lookup search algorithm starts to be
efficient, and most of the queries are resolved without the need for
trie-search.
Finally, we show the scalability of starcode with increasing
number of threads in Figure~\ref{fig:perf}d. The search algorithm is fully
parallel and the relative performance increases linearly up to 8 threads.
The bending observed thereafter has two sources. The first is 
that the input reading and clustering steps are brief but not parallel,
the second is that there is insufficient memory bandwidth to satisfy
the increased demand of memory accesses due to hardware limitations.

@@

We generated 1 million random 40-mers duplicated 47 times. For each
40-mer, we also inserted 3 additional mutant sequences to the pool
by sampling again 3 nucleotides at random positions. The probability
that such a mutant sequence is within a distance 3 of any other
40-mer is of the order of $10^{-12}$ and can be neglected. Each
cluster thus consists of 50 sequences, of which 3 have a Levenshtein
distance 3 from the canonical representant (referred to as the centroid
in clustering terms).

All software were set to run on a 6-core single-processor Intel Xeon
E5-2620 system with 48~GB of DDR3-RAM at 1333~Mhz.
Using 12 threads, starcode clustered the shuffled 50 million sequences
in 5 minutes and 3 seconds without a single error (the output
consisted of 1 million clusters of size 50). There is no obvious
way to compare exact algorithms with heuristic algorithms such as
CD-HIT and USEARCH. We decided to allocate the heuristics 6 minutes
on the same dataset (20\% additional time) and use the
parameters giving the best accurary under this constraint.
However, we could not find a combination of
parameters such that either of them could terminate in less than
6 minutes. To simplify the task, we extracted the unique sequences
from the input file, bringing the number down to 4,000,000 and followed
the same rationale. We could not find any set of parameters such
that USEARCH could process this input in less than 6 minutes
(because of poor multi-threading performance), but CD-HIT could
perform the task when run as shown in Table~\ref{tab:params1}.
In such conditions, it identified 2,798,656 clusters, indicating that
the clustering was not nearly complete.

\begin{table}[h]
\centering
\caption{Software execution parameters (benchmark 1)}
\resizebox{\columnwidth}{!}{
\begin{tabular}{|l|l|}
    \hline
    starcode & \texttt{-d3 -t12 -s}\\ \hline
    CD-HIT 4.6.1 & \texttt{cd-hit-est -n 7 -c .925 -T 12 -M 0 -r 0 -A 40}\\ \hline
\end{tabular}
}
\label{tab:params1}
\end{table}

We next benchmarked starcode on the biological dataset used by
\cite{pmid23060610}. Because USEARCH imposes a memory limit on the
size of the input file, we clustered the forward reads on 95\%
identity from one lane only (reference ERR011089), adding up to
4,189,859 reads, each 75 nucleotides long. Using 12 threads, starcode
identified 4,111,111 clusters in 6 minute 19 seconds. Allocating up to
7 minutes 30 seconds to CD-HIT and USEARCH, we selected the parameters
giving highest accuracy. In such conditions (see Table~\ref{tab:params2})
CD-HIT identified 4,126,352 clusters while USEARCH could not be run in
the allocated time. These results indicate that starcode achieves good
performance on clustering high throughput sequencing reads.

\begin{table}[h]
\centering
\caption{Software execution parameters (benchmark 2)}
\resizebox{\columnwidth}{!}{
\begin{tabular}{|l|l|}
    \hline
    starcode & \texttt{-d4 -t12 -s}\\ \hline
    CD-HIT 4.6.1 & \texttt{cd-hit-est -n 8 -c .95 -T 12 -M 0 -r 0 -A 75}\\ \hline
\end{tabular}
}
\label{tab:params2}
\end{table}

\begin{figure}[!tpb]
\centerline{\includegraphics{benchmark.pdf}}
\caption{Benchmark of starcode against short read mappers.
\textbf{a} Running time compared to Bowtie2, BWA and GEM on the
dataset described in the text. The running time is decomposed as
extraction of unique sequences (purple), Burrows-Wheeler indexing
(yellow) and query proper (white). The extraction was performed with
the Linux \texttt{sort} command, so the time is the same in every
case. \textbf{d} Error rate compared to Bowtie2, BWA and GEM,
expressed as the proportion of matches with a single hit. This is an
underestimate of the true error rate.}
\label{fig:benchmark}
\end{figure}

\subsection{Benchmark against short read mappers}
A potential strategy to cluster short sequences is to use
Burrows-Wheeler transform-based short read mappers to match a set of
sequences against itself. Once matches are available, several community
detection algorithms can be used to identify the clusters. We 
benchmarked starcode against the short read mappers Bowtie2
\citep{pmid22388286}, BWA \citep{pmid19451168} and GEM
\citep{pmid23103880}. Short read mappers do not perform clustering, so
we only evaluated their performance on the matching problem.

For this test, we used the first dataset described above, in which
every sequence has at least 2 distinct matches at a distance 3 (mutant
sequences from the same cluster may be at a distance up to 6 from each
other, but their distance to the centroid is always 3). This means that
we can use matches with a single hit (every read trivially matches
itself) as a lower bound on the inaccuracy. We used the 4,000,000 unique
40-mers to build an index that was later queried with the same 40-mers
using parameters shown in Table~\ref{tab:srm}.

\begin{table}[h]
\centering
\caption{Software execution parameters (benchmark 3)}
\resizebox{\columnwidth}{!}{
\begin{tabular}{|l|l|}
    \hline
    Bowtie2 indexer & \texttt{bowtie2-build --quiet}\\ \hline
    Bowtie2 2.1.0 & \texttt{-f -a -p 12 --no-hd --very-sensitive}\\ \hline
    BWA indexer & \texttt{bwa index -a is}\\ \hline
    BWA 0.7.9a & \texttt{mem -t 12 -a -k1 -B0} \\ \hline
    GEM indexer & \texttt{gem-indexer -T 12} \\ \hline
    GEM 1.423 & \texttt{-e3 -s4 -T12 --granularity 100000}\\ \hline
\end{tabular}
}
\label{tab:srm}
\end{table}

Figure~\ref{fig:benchmark}a shows the running time of starcode along
with the decomposition of the running time for the three mappers. Note
that the time to extract unique sequences is always the same because
we used the Linux \texttt{sort -u} command in every case. The running
time of starcode is slightly lower than short read mappers when all
steps ares taken into account. Figure~\ref{fig:benchmark}b shows the
lower bound on the error rate, as estimated by the proportion of single
hits. BWA was found to be the most accurate aligner with a lower bound
of 31\% sequences with a single match. This measure is an underestimate
of the true error rate, but given that starcode is an exact algorithm,
a more accurate measure is not necessary: starcode can compete with the
short read aligners on the matching problem, and it provides a precision
that none of the other tools can offer.

@@

We tested the efficiency of starcode on the TRIP dataset from
\cite{pmid23953119}. In the experiment labeled mPGKA, we identified
24.1 million (91\%) barcode-containing reads out of 26.6 million,
consisting of approximately 223,000 and 220,000 unique barcode sequences
for PCR replicates 1 and 2, respectively. Following the authors, we kept
only the barcodes with at least 5 counts and performed clustering as
described: ``First we sorted barcodes according to their counts. Then,
for each barcode (starting from the most frequent one), we identified
and removed all its mutant versions, defined as barcodes within a Hamming
distance of 2.'' We implemented this method, here on referred to as the
``sequential algorithm'' with the function call
\texttt{stringDist(method='lv', maxDist=2)} from R the package
\texttt{stringDist} \citep{R}.  When replacing the Hamming distance by
the Levenshtein
distance, starcode produced exactly the same output as the sequential
algorithm. The running time of starcode averaged over the replicates was
2.90 seconds, \textit{versus} more than 3 minutes 40 seconds, which
represents a 75-fold speedup.

@@

We set up a test based on the meningitis-causing agent
\textit{Neisseria meningitidis}. The genome of this bacterium is
interspersed with a frequent 12 bp sequence known as DNA uptake
sequence \citep{pmid10673000}. We extracted the 12-mers from both
orientations of  the 2.19 Mb genome, yielding 4.39 million 12-mers,
consisting of 2.77 million unique sequences. Clustering the 12-mers
with starcode within a Levenshtein distance of 2 took less than 45
seconds with 12 threads. We identified the known DNA uptake sequence
of \textit{Neisseria meningitidis} (ATGCCGTCTGAA) as the most abundant
12-mer, with 1466 exact and 2096 inexact hits. This result testifies
to the fact that starcode can be used to identify biologically
relevant motifs in bacterial genomes.

To test starcode on another application, we used the RNA-protein
interaction data produced by RNAcompete \citep{pmid19561594}. The
mammalian splicing factor SRSF1 is known to bind RNA GA-rich motifs,
but there is some disagreement about the motif that it recognizes
\citep{pmid23562324}. For each replicate of the human SRSF1 in the
RNAcompete dataset, we replaced the microarray signals by their rank
and extracted the 10-mers from the microarray probes. The 10-mers were
given a score equal to the rank of the probe they belong, and enriched
motifs were found using the sphere clustering of starcode with maximum
Levenshtein distance 2. The score of the most enriched 10-mer is thus
the sum of the ranks of all 10-mers within this distance. Clustering
the 6.3 million extracted 10-mers with 12 threads took about 20 seconds
for each replicate.  The most enriched 10-mers were AGGACACGGA,
AGGACACGGA, AGGACGGAGG, AGGACGGAGG, AGGACACGGA and AGGATACAGG. Except
for the last replicate, the motifs consist of AGGAC and GGA, with a
spacer of variable length. This suggests that the binding of SRSF1 to
RNA may involve a spacer sequence, which would explain the disagreement
between the motifs derived from 6-mers or 7-mers.

\section{Discussion and conclusion}
Through the parallel poucet search algorithm, starcode implements an
exact sequence clustering algorithm that can be faster than popular
heuristics. By design, starcode is tailored to process high throughput
sequencing data on multi-core platforms. Our benchmark shows that
starcode achieves perfect clustering on short random sequences in less
time than what is considered acceptable for heuristic searches. We also
show that starcode outperforms next generation read mappers in this
context. The software tools used for this benchmark are usually not
used for barcode or random sequence clustering. In this respect,
starcode fills a need arising from the development of barcoding
technologies.

The speed of starcode also makes it useful for other clustering
tasks, such as identifying enriched motifs in microbial genomes and
in experimental data. Here we have given two examples of such
applications. In the first, we recover a known enriched 12-mer in the
genome of \textit{Neisseria meningitidis}. In the second, we recover
the motif of the human RNA binding protein SRSF1 and notice that it
seems to consist of two halves separated by a linker. This hypothesis
is consistent with the fact that SRSF1 binds RNA through two
consecutive RNA Recognition Motifs (RRM) that are known to bind
3-4 nucleotides in a row\citep{pmid23253355}. The Levenshtein distance,
which incorporates insertions and deletions is more likely to capture
bi-partite binding motifs than position weight matrix representations.
The use of a clustering method to tackle this problem is unusual, but it 
it illustrates the potential advantages of distance-based approaches.

The current version of starcode has been primarily optimized for speed.
The memory footprint depends on the number of sequences to cluster
(because the sequences of the input set are loaded in memory as a set of
of tries) and on the mean number of matches per sequence. Every match is
stored in memory until the clustering phase, which may represent a large
overhead if the dataset is dense. As counterintuitive as it may seem,
long queries will usually impose a lower memory footprint because the
matches between the sequences are more sparse. We have shown that the
running time will also be shorter thanks to the lookup table
search (Figure~\ref{fig:perf}c).

Perhaps the most surprising element of this study is that an exact
algorithm can compete with extremely fast heuristics. This will no
longer be true when clustering divergent sequences because the Levenshtein
distance will have to be increased, leading to exponentially longer
running times (Figure~\ref{fig:perf}b). However, for the imporant
practical case that the divergence is driven by sequencing errors,
starcode illustrates that there is still room for algorithmic innovations
that can outperform heuristics. The idea of the poucet search seems
simple in retrospect, yet it is a powerful way to tap into the data
structuration provided by string sorting. This principle could find some
applications in algorithms used in various fields.

@@

\section*{Acknowledgement}
We would like to thank Maria Chatzou for her precious feedback on the
preliminary version of this manuscript.
